Longitudinal Principal Manifold Estimation

### 2023-07-07 paper/main.tex
update application section

To solve this problem, we propose an extension of the PME algorithm developed in \cite{mengPrincipalManifoldEstimation2021}. The output of the PME algorithm run used for the data at an individual time point is a spline function mapping from a low-dimensional paramterization to a high-dimensional space. In the LPME algorithm, we use the coefficients of PME functions fit at each time point as the outcomes in a thin plate spline, thus smoothing over the coefficients with respect to time. This yields a multistage model that, for a provided time point and set of parameters, obtains a set of spline coefficients associated with the time value, then returns the high-dimensional output of the spline function associated with the parameter set. Our proposed LPME algorithm consists of four steps: data reduction, initialization, fitting, and tuning. Details of this approach are described below.

@@

While the original study included 200 cognitively healthy elderly individuals, 400 with mild cognitive impairment, and 200 with AD, for the purposes of demonstration we used $T_1$-weighted structural MRI images from eight study participants in the cognitively normal group (\cite{jack2008adni}). Follow-up duration for these participants ranged between approximately 16 months to 48 months, with imaging scheduled to be conducted at six month intervals. Individuals from the cognitively normal group were selected because drastic changes in the volume and shape of subcortical structures over the course of the study follow-up duration would be unexpected. Under this assumption, we attribute volumetric changes in a given structure to error introduced by the study procedure or the image segmentation process. Due to the changes experienced by those with Alzheimer's disease, we focused on the hippocampi of the selected individuals. Images were processed using FSL and ANTs via the \texttt{fslr} and \texttt{ANTsR} packages in \texttt{R}. Specifically, ANTs was used for bias correcting the images, and the FIRST method in FSL was used for image segmentation.

>>>>

To solve this problem, we propose an extension of the PME algorithm developed in \cite{mengPrincipalManifoldEstimation2021}. The output of the PME algorithm run used for the data at an individual time point is a spline function mapping from a low-dimensional paramterization to a high-dimensional space. In the LPME algorithm, we use the coefficients of PME functions fit at each time point as the outcomes in a thin plate spline, thus smoothing over the coefficients with respect to time. This yields a multistage model that, for a provided time point and set of parameters, obtains a set of spline coefficients associated with the time value, then returns the high-dimensional output of the spline function associated with the parameter set. Our proposed LPME algorithm consists of four steps: data reduction, initialization, fitting, and tuning. Details of this approach are described below, and an implementation is available as an \texttt{R} package at \texttt{https://github.com/rjzielinski/pme}.

@@

While the original study included 200 cognitively healthy elderly individuals, 400 with mild cognitive impairment, and 200 with AD, for the purposes of demonstration we used $T_1$-weighted structural MRI images from eight study participants in the cognitively normal group (\cite{jack2008adni}). Follow-up duration for these participants ranged between approximately 16 months to 48 months, with imaging scheduled to be conducted at six month intervals. Individuals from the cognitively normal group were selected because drastic changes in the volume and shape of subcortical structures over the course of the study follow-up duration would be unexpected. Under this assumption, we attribute volumetric changes in a given structure to error introduced by the study procedure or the image segmentation process. Due to the changes experienced by those with Alzheimer's disease, we focused on the hippocampi of the selected individuals.

Images were processed using FSL and ANTs via the \texttt{fslr} and \texttt{ANTsR} packages in \texttt{R}. Specifically, ANTs was used for bias correcting the images, and the FIRST method in FSL was used for image segmentation. Following image segmentation, the surfaces of each hippocampus were identified by finding the extreme voxels in each dimension with nonzero intensity readings. The estimated hippocampus surface positions were then standardized to a maximum distance of one from the origin in Euclidean space and centered around the origin.

Provide a brief summary of the results of the model fitting here, along with the related figures.
